---
title: "Mutation base spectrum"
author: Ahmed Hasan
output: pdf_document
---

# Package import

```{r}
library(tidyverse)
library(fs)
library(here)
library(wesanderson)
```

# Data import

```{r}
# all mut describer outputs
salt <- read_tsv(
  here('data/rate/mut_describer/muts_described.final.gene_sets.tsv'), col_types = cols()) %>% 
  mutate(sample_type = str_extract(mutant_sample, '[05]$'),
         sample_group = str_extract(mutant_sample, '^[A-Z]{2}')) %>% 
  mutate(sample_factor = paste(sample_group, sample_type, sep = '_'))

ma <- read_tsv(
  here('data/spectrum_context/MA/final.curated_muts.coord_sorted.txt'), 
  col_types = cols(), na = c('.')
) %>% 
  select(-chromosome_1, -position_1) %>% 
  filter(nchar(mutation) == 3) # remove indels

adaptation <- read_csv(
  here('data/prev/all_mutations_w_shared_hmmIBD_corrected_FPKM.csv'), 
  col_types = cols(), na = c('.')
) %>% 
  filter(type == 'SNP')

```

# Mutational base spectrum

6 possible SNMs
- A -> C (also T -> G)
- A -> G (also T -> C)
- A -> T (also T -> A)
- C -> A (also G -> T)
- C -> G (also G -> C)
- C -> T (also G -> A)

## Generating observed count tables

Need to get splice sites for silent/nonsilent classification of intronic sites - 
ran this on hpcnode1 using a gffutils db:

```{python, eval=FALSE}
import csv
import gffutils
from tqdm import tqdm

with open('muts.csv', 'r') as f: # equivalent of adaptation data frame above
    reader = csv.DictReader(f)
    reclist = [rec for rec in reader if rec['intronic'] == '1']
  
db = gffutils.create_db('final.strict.GFF3', dbfn='final.strict.db',
    force=True, keep_order=True, merge_strategy='merge', sort_attribute_values=True)
db = gffutils.FeatureDB('final.strict.db', keep_order=True)

splice_list = []
for record in tqdm(reclist):
    gene_name = record['Gene.primaryIdentifier']
    db_gene = db[gene_name]
    pos = int(record['position'])
    for exon in db.children(db_gene, featuretype='exon', order_by='start'):
        if min(abs(pos - exon.start), abs(pos - exon.end)) < 3:
        print(gene_name, record['chromosome'], record['position'])
        print(exon)
        splice_list.append(record)
```

Got me four intronic mutations that shouldn't be classed as silent:

```
Cre12.g545550 chromosome_12 8771346
chromosome_12	phytozome8_0	exon	8770962	8771344	.	+	.	ID=PAC:26907314.exon.5;pacid=26907314;Parent=PAC:26907314;ness_ID=26907314
g5391 chromosome_5 2315498
chromosome_5	phytozome8_0	exon	2315470	2315497	.	-	.	ID=PAC:26897662.exon.6;pacid=26897662;Parent=PAC:26897662;ness_ID=26897662
Cre06.g265550 chromosome_6 2122297
chromosome_6	phytozome8_0	exon	2122193	2122295	.	-	.	ID=PAC:26893506.exon.4;pacid=26893506;Parent=PAC:26893506;ness_ID=26893506
Cre08.g362650 chromosome_8 1131421
chromosome_8	phytozome8_0	exon	1131423	1131551	.	-	.	ID=PAC:26888200.exon.20;pacid=26888200;Parent=PAC:26888200;ness_ID=26888200
```

Creating a data frame for these:

```{r}
splice_sites = data.frame(
  chromosome = c('chromosome_12', 'chromosome_5', 'chromosome_6', 'chromosome_8'),
  position = c(8771346, 2315498, 2122297, 1131421)
)

# adding a silent/nonsilent column to the adaptation data frame
adaptation_silent_nonsilent = adaptation %>% 
  mutate(site_type = case_when(
    Nonsyn_V_Syn != 'synonymous' & intronic == 0 & intergenic == 0 ~ 'nonsilent',
    Nonsyn_V_Syn == 'synonymous' | intronic == 1 | intergenic == 1 ~ 'silent'
  )) %>% 
  mutate(site_type = ifelse(
    chromosome %in% splice_sites$chromosome & position %in% splice_sites$position, 'nonsilent', site_type
  )) # 119 each

```

Getting mutation counts:

```{r}
get_counts = function(d) {
  count_table <- d %>% 
    select(mutation) %>% 
    mutate(
      type = case_when(
        mutation == 'T>G' ~ 'A>C',
        mutation == 'T>C' ~ 'A>G',
        mutation == 'T>A' ~ 'A>T',
        mutation == 'G>T' ~ 'C>A',
        mutation == 'G>C' ~ 'C>G',
        mutation == 'G>A' ~ 'C>T',
        TRUE ~ mutation
      )
    ) %>% 
    count(type, name = 'count')
  return(count_table)
}

count_table = list(
    # salt_nonsilent = salt %>% filter(nonsyn_v_syn != 'nonsynonymous'), 
    # salt_silent = salt %>% filter(nonsyn_v_syn == 'synonymous'),
    salt_0 = salt %>% filter(sample_type == '0'),
    salt_5 = salt %>% filter(sample_type == '5'),
    ma = ma, 
    # adaptation = adaptation
    adaptation_nonsilent = adaptation_silent_nonsilent %>% filter(site_type == 'nonsilent'),
    adaptation_silent = adaptation_silent_nonsilent %>% filter(site_type == 'silent')
  ) %>% 
  map_dfr(., get_counts, .id = 'name')

# separating three salt MA groups in separate count table
salt_group_table = salt %>% 
  mutate(group_type = paste(sample_group, sample_type, sep = '_')) %>% 
  split(.$group_type) %>% 
  map_dfr(., get_counts, .id = 'name')
  
count_table

```

## Calculate expected mutation counts

Expected mutation rates:

```{r}
obs_exp = count_table %>% 
  # get the total mutations per anc mut per sample (e.g. all A anc muts in saltMA)
  mutate(anc_mut = substr(type, 1, 1)) %>% 
  mutate(group = str_extract(name, '(salt_[05]|ma|adaptation)')) %>% 
  group_by(name) %>% 
  mutate(total_mut_anc = sum(count)) %>% 
  ungroup() %>% 
  # get expected number of muts (1/3 prob for either base)
  mutate(expected_count = case_when(
    anc_mut == 'A' ~ ceiling((total_mut_anc * 0.365) / 3),
    anc_mut == 'C' ~ ceiling((total_mut_anc * 0.635) / 3))
    ) %>% 
  # get obs / exp
  mutate(obs_exp = count / expected_count)

obs_exp

obs_exp_salt = salt_group_table %>% 
  mutate(anc_mut = substr(type, 1, 1)) %>% 
  group_by(name) %>% 
  mutate(total_mut_anc = sum(count)) %>% 
  ungroup() %>% 
  # get expected number of muts
  mutate(expected_count = case_when(
    anc_mut == 'A' ~ ceiling((total_mut_anc * 0.365) / 3),
    anc_mut == 'C' ~ ceiling((total_mut_anc * 0.635) / 3))
    ) %>% 
  # get obs / exp
  mutate(obs_exp = count / expected_count) %>% 
  # add missing rows
  add_row(name = 'CC_0', type = 'A>T') %>% 
  add_row(name = 'SL_0', type = 'C>G') %>% 
  add_row(name = 'SL_0', type = 'C>T') %>% 
  add_row(name = 'SL_5', type = 'A>T') %>% 
  add_row(name = 'SL_5', type = 'A>G') %>% 
  add_row(name = 'SL_5', type = 'C>T') %>% 
  add_row(name = 'SL_5', type = 'C>G') %>% 
  mutate(across(where(is.numeric), .fns = ~ ifelse(is.na(.), 0, .))) %>% 
  mutate(anc_mut = substr(type, 1, 1))

obs_exp_salt
  
```

## Plots

```{r}
obs_exp %>% 
  mutate(
    xcol = str_replace(type, '>', ' \u2192 '),
    name = case_when(
      name == 'adaptation_nonsilent' ~ 'HS lines (non-silent sites)',
      name == 'adaptation_silent' ~ 'HS lines (silent sites)',
      name == 'ma' ~ 'MA',
      name == 'salt_0' ~ 'Salt MA (control)',
      name == 'salt_5' ~ 'Salt MA (salt)'
    )) %>% 
  mutate(
    name = fct_relevel(name, 
      c('MA', 'HS lines (non-silent sites)', 'HS lines (silent sites)', 
        'Salt MA (control)', 'Salt MA (salt)'))
  ) %>% 
ggplot(
  aes(x = xcol, y = obs_exp, fill = name)) +
  geom_bar(
    stat = 'identity', 
    position = position_dodge(),
    color = 'black') +
  geom_hline(aes(yintercept = 1), linetype = 'dashed') +
  coord_cartesian(
    y = c(0, 2.5)
  ) +
  theme_classic() +
  labs(
    x = 'Base change type',
    y = 'Observed over expected'
  ) +
  theme(
    axis.text = element_text(size = 12, color = 'black'),
    axis.title = element_text(size = 12, color = 'black'),
    legend.title = element_blank(),
    # legend.position = c(0.2, 0.88),
    legend.position = 'top',
    legend.background = element_blank(),
    legend.text = element_text(size = 10, color = 'black')
  ) +
  guides(
    fill = guide_legend(ncol = 2, nrow = 3)
  ) +
  scale_fill_manual(values = wes_palette(7, name = 'Darjeeling1', type = 'continuous')[3:7])


```

Separating the MA/HS and individual groups for salt MA:

```{r}
palette = wes_palette(12, name = 'Darjeeling1', type = 'continuous')

obs_exp_salt %>% 
  mutate(
    xcol = str_replace(type, '>', ' \u2192 ')) %>% 
ggplot(
  aes(x = xcol, y = obs_exp, fill = name)) +
  geom_bar(
    stat = 'identity', 
    position = position_dodge(),
    color = 'black') +
  geom_hline(aes(yintercept = 1), linetype = 'dashed') +
  coord_cartesian(
    y = c(0, 2.5)
  ) +
  theme_classic() +
  labs(
    x = 'Base change type',
    y = 'Observed over expected'
  ) +
  theme(
    axis.text = element_text(size = 12, color = 'black'),
    axis.title = element_text(size = 12, color = 'black'),
    legend.title = element_blank(),
    # legend.position = c(0.2, 0.88),
    legend.position = 'top',
    legend.background = element_blank(),
    legend.text = element_text(size = 10, color = 'black')
  ) +
  guides(
    fill = guide_legend(ncol = 3, nrow = 2)
  ) +
  scale_fill_manual(
    values = c(palette[4], palette[5], # CC
               palette[7], palette[9], # DL 
               palette[11], palette[12])) # SL

```

# Equilibrium GC content

$$ \frac{n_{AT -> GC}}{n_{GC -> AT} + n_{AT -> GC}} $$

where $n_{AT -> GC}$ is a 'W->S' mutation, and the reverse is a 'S->W' mutation 
(Mugal, Weber, Ellegren 2015)

```{r}
# GC 68891655 
# AT 38724292
# total 107615947

# GC 37413853 
# AT 20922472
# total 58336325

get_eq_gc = function(d, silent = FALSE, values = FALSE) {
  if (!silent) {
    num_GC = 68891655
    num_AT = 38724292
  } else if (silent) {
    num_GC = 37413853
    num_AT = 20922472
  }
  ws_mutations = c('A>G', 'A>C', 'T>G', 'T>C')
  sw_mutations = c('G>A', 'G>T', 'C>A', 'C>T')
  
  n_ws = d %>% 
    filter(mutation %in% ws_mutations) %>% 
    nrow()
  
  n_sw = d %>% 
    filter(mutation %in% sw_mutations) %>% 
    nrow()
  
  n_ws_norm = n_ws / num_AT
  n_sw_norm = n_sw / num_GC
  
  if (values) {
    return(c(
      'n_ws' = n_ws,
      'n_sw' = n_sw,
      'num_AT' = num_AT,
      'num_GC' = num_GC))
  } else if (!values) {
    eq_gc = n_ws_norm / (n_ws_norm + n_sw_norm)
    message(paste(n_ws, n_sw))
    message(paste(n_ws / num_AT, n_sw / num_GC))
    return(eq_gc)
  }
}
```

Calculate values:

```{r}
get_eq_gc(ma)

get_eq_gc(adaptation_silent_nonsilent %>% filter(site_type == 'silent'), silent = TRUE)

map(
  salt %>% split(.$sample_factor),
  ~ get_eq_gc(.)
)
```



## WS and SW mutation counts

```{r}
# MA orig - mutations to ATs vs mutations to GCs
ma_orig_values = get_eq_gc(ma, values = TRUE)
message('ma')
as.numeric(ma_orig_values['n_sw'] / ma_orig_values['n_ws']) # 4.65

# adaptation - silent and nonsilent
adaptation_silent_values = get_eq_gc(
  adaptation_silent_nonsilent %>% 
    filter(site_type == 'silent'), 
  values = TRUE)
adaptation_nonsilent_values = get_eq_gc(
  adaptation_silent_nonsilent %>% 
    filter(site_type == 'nonsilent'), 
  values = TRUE)

message('adaptation silent')
as.numeric(adaptation_silent_values['n_sw'] / adaptation_silent_values['n_ws']) # silent - 3.65
message('adaptation nonsilent')
as.numeric(adaptation_nonsilent_values['n_sw'] / adaptation_silent_values['n_ws']) # nonsilent - 3.75

# salt ma - across all factors
map(
  salt %>% split(.$sample_factor),
  function(d) {
    vals = get_eq_gc(d, values = TRUE)
    mut_ratio = vals['n_sw'] / vals['n_ws']
    return(as.numeric(mut_ratio))
  }
)


```



# Ts/Tv

```{r}
ts_tv = function(d, values = FALSE) {
  tv = c('A>T', 'T>A', 'G>C', 'C>G', 'A>C', 'C>A', 'G>T', 'T>G')
  ts = c('A>G', 'G>A', 'C>T', 'T>C')
  ts_count = d %>% 
    filter(mutation %in% ts) %>% 
    nrow()
  tv_count = d %>% 
    filter(mutation %in% tv) %>% 
    nrow()
  ts_tv_ratio = ts_count / tv_count
  if (!values) {
    return(ts_tv_ratio)
  } else if (values) {
    return(c('ts' = ts_count, 'tv' = tv_count))
  }
}
```

Running over datasets:

```{r}
message('ma')
ts_tv(ma)

message('adaptation')
ts_tv(adaptation_silent_nonsilent)
ts_tv(adaptation_silent_nonsilent %>% filter(site_type == 'silent'))
ts_tv(adaptation_silent_nonsilent %>% filter(site_type == 'nonsilent'))

message('salt_ma')
map(
  salt %>% 
    split(.$sample_factor),
  ~ ts_tv(.)
)
```

Binomial tests - classifying Ts as 'success', meaning prob(success) is 1/3 -

```{r}
# original MA
ma_vals = ts_tv(ma, values = TRUE)
binom.test(ma_vals['ts'], sum(ma_vals), alternative = 'two.sided', conf.level = 0.95) 

d_list = list(
  'ma' = ma, 
  'adaptation' = adaptation_silent_nonsilent,
  'adaptation_silent' = adaptation_silent_nonsilent %>% filter(site_type == 'silent'),
  'adaptation_nonsilent' = adaptation_silent_nonsilent %>% filter(site_type == 'nonsilent'))

salt_list = salt %>% split(.$sample_factor)

d_list = append(d_list, salt_list)

map_dfr(
  d_list,
  function(d) {
    vals = ts_tv(d, values = TRUE)
    out = binom.test(
      vals['ts'], sum(vals), p = 1/3,
      alternative = 'two.sided',
      conf.level = 0.95
    ) %>% 
      broom::tidy()
  },
  .id = 'sample_factor'
)


```
































