---
title: "KaKs"
author: Ahmed Hasan
output: pdf_notebook
---

# Package import

```{r}
library(tidyverse)
library(fs)
library(here)
library(boot)
library(broom)
library(tictoc)
```

# Data import

```{r}
all_genome = read_tsv(here('data/rate/ka_ks/all_genome.tsv'), col_types = cols()) %>% 
  mutate(sample_type = case_when(
    str_ends(sample, '0') ~ 0,
    str_ends(sample, '5') ~ 5
  )) %>% 
  mutate(sample_group = str_extract(sample, '^[A-Z]{2}'))

all_genome_callables = read_tsv(here('data/rate/ka_ks/degen_callables_lookup.tsv'), col_types = cols()) %>% 
  mutate(sample_type = case_when(
    str_ends(sample, '0') ~ 0,
    str_ends(sample, '5') ~ 5
  )) %>% 
  mutate(sample_group = str_extract(sample, '^[A-Z]{2}'))

all_perrineau = read_tsv(
  here('data/rate/ka_ks/syn_nonsyn_perrineau_counts.tsv'), col_types = cols()) %>% 
  mutate(total_muts = rowSums(
    across(contains('muts')))
  ) %>% 
  filter(total_muts > 0) %>%
  select(-total_muts) 
  
all_perrineau_callables = read_tsv(
  here('data/rate/ka_ks/perrineau_genes_callables.tsv'), col_types = cols())

all_expressed = read_tsv(
  here('data/rate/ka_ks/syn_nonsyn_expressed_counts.tsv'), col_types = cols()) %>% 
  mutate(total_muts = rowSums(
    across(contains('muts')))
  ) %>% 
  filter(total_muts > 0) %>%  # 41 mutated genes
  select(-total_muts) 
  
all_expressed_callables = read_tsv(
  here('data/rate/ka_ks/expressed_genes_callables.tsv'), col_types = cols())
```

# Ka/Ks calc

## Genomewide

Need to calculate:

$$

\frac{K_a}{K_s} = \frac{\text{nonsyn muts} / \text{nonsyn callables}}{\text{syn muts} / \text{syn callables}}

$$

```{r}
# get callables per sample type
genome_sample_callables = all_genome_callables %>% 
  # filter(str_detect(scaffold, 'scaffold', negate = TRUE)) %>% 
  group_by(sample_group, sample_type) %>% 
  summarise(across(contains('syn'), sum))

# get muts per sample type
genome_sample_counts = all_genome %>% 
  select(-nonsyn, -syn) %>% # these values only consider chrs with muts
  group_by(sample_group, sample_type) %>% 
  summarise(across(.cols = contains('syn'), .fns = sum))
```

and now to calculate Ka/Ks:

```{r}
ka_ks_genome = genome_sample_counts %>% 
  
# join callables and counts
  full_join(
    genome_sample_callables, by = c('sample_group', 'sample_type')
  ) %>% 
  replace_na(list(nonsyn_muts = 0, syn_muts = 0)) %>% # SL5 has none

# compute Ka, Ks, and Ka/Ks values for 0 and 5
  mutate(
    ka = nonsyn_muts / nonsyn,
    ks = syn_muts / syn
  ) %>% 
  mutate(
    ka_ks = ka / ks
  ) %>% 
  select(sample_group, sample_type, ka_ks, everything())

ka_ks_genome
```

## Bootstrap confidence intervals

Need to bootstrap original mutation dataset for this -

### Load in and prepare data

```{r}
muts_described = read_tsv(
  here('data/mutations/mut_describer/muts_described.final.tsv'), 
  col_types = cols()
) %>% 
  
  # only keep cols of interest and remove sites not classed as syn/nonsyn
  select(
    chromosome, position, mutant_sample, nonsyn_v_syn
  ) %>% 
  filter(!is.na(nonsyn_v_syn)) %>% 
  
  # add sample group and type as cols
  mutate(
    sample_group = str_extract(mutant_sample, '^[A-Z]{2}'),
    sample_type = str_extract(mutant_sample, '[05]$')
  )
```

### Resampling muts with replacement

```{r}
ka_ks_calc = function(d, indices) {
  d_resampled = d[indices,]
  d_summarised = d_resampled %>% 
    
    # restructure resampled dataset to match join from earlier
    group_by(sample_group, sample_type) %>% 
    count(nonsyn_v_syn) %>% 
    ungroup() %>% 
    pivot_wider(names_from = nonsyn_v_syn, values_from = n, values_fill = 0) 
  
  if (!'synonymous' %in% colnames(d_summarised)) {
    d_summarised$synonymous = 0
  }
  
  # get sample group and type
  current_sample_group = unique(d_summarised$sample_group)
  current_sample_type = unique(d_summarised$sample_type)
  
  # get callables to compute ka/ks
  d_summarised = d_summarised %>% 
    rename(nonsyn_muts = nonsynonymous, syn_muts = synonymous) %>% 
    mutate(sample_type = as.numeric(sample_type)) %>% 
    
    # get ka/ks
    full_join(
      filter(genome_sample_callables,
             sample_group == current_sample_group,
             sample_type == current_sample_type), 
      by = c('sample_group', 'sample_type')
    ) %>% 
    replace_na(
      list(nonsyn_muts = 0, syn_muts = 0)
    ) %>% 
    mutate(
      ka = nonsyn_muts / nonsyn,
      ks = syn_muts / syn
    ) %>%
    mutate(
      ka_ks = ka / ks) %>% 
    
    # remove extra columns and prep for 'export'
    select(sample_group, sample_type, ka_ks) %>% 
    ungroup()

    return(d_summarised$ka_ks)
}

```

### Calculate confidence intervals from bootstrap vals

```{r}
get_cis = function(d, rep_count) {
  tic(msg = paste('bootstrap for', unique(d$sample_factor)))
  message(paste('starting', unique(d$sample_factor)))
  boot_obj = boot(data = d, statistic = ka_ks_calc, R = rep_count)
  ci_obj = boot.ci(boot_obj, type = 'bca')$bca[c(4,5)] # lower and upper bounds
  out = data.frame(
    sample_group = unique(d$sample_group),
    sample_type = unique(d$sample_type),
    lower_ci = ci_obj[1],
    upper_ci = ci_obj[2]
  )
  toc(log = TRUE)
  return(out)
}

ka_ks_cis = muts_described %>% 
  mutate(sample_factor = paste(sample_group, sample_type, sep = '_')) %>% 
  filter(sample_group != 'SL') %>% 
  split(.$sample_factor) %>% 
  map_dfr(~ get_cis(., 1000)
  ) 

ka_ks_cis

```

### Combining with Ka/Ks values above

```{r}
ka_ks_genome_final = ka_ks_genome %>% 
  mutate(sample_type = as.character(sample_type)) %>% 
  full_join(
    ka_ks_cis,
    by = c('sample_group', 'sample_type')) %>% 
  select(sample_group, sample_type, ka_ks, contains('ci'), everything())

ka_ks_genome_final

write_tsv(
  ka_ks_genome_final, 
  file = here('data/rate/ka_ks/ka_ks_genome_final_CIs.tsv'))
```



## Expressed genes

How many genes had mutations in multiple samples? 

```{r}
all_expressed %>% 
  group_by(gene_name) %>% 
  count() %>% 
  arrange(-n) # just NP_042564.1
```

Creating a gene name column:

```{r}
all_genes = unique(all_expressed$gene_name)
all_gene_names = all_expressed_callables$gene_name
g_temp = character(length(all_gene_names))

get_gene = function(string, pattern) {
  gene_vec = str_extract(string, pattern)
  gene_name = gene_vec[!is.na(gene_vec)]
  return(gene_name)
}

# pain
# no idea why mutate() refused to work for this operation
for (i in seq_along(g_temp)) {
  gene = get_gene(all_gene_names[i], all_genes)
  g_temp[i] = gene
}

all_expressed_callables$actual_gene_name = g_temp
```

Joining the callables dataset with the mutation counts and computing Ka/Ks:

```{r}
expressed_ka_ks = all_expressed %>% 
  
  # transform to long format and separate by sample type
  pivot_longer(cols = contains('muts'), names_to = 'measure') %>% 
  mutate(
    sample_group = str_extract(sample, '^[A-Z]{2}'),
    sample_type = str_extract(sample, '([05])$'), 
    measure_type = str_extract(measure, '([0-5])$')) %>% 
  mutate( 
    measure = str_replace(measure, '_[05]', '')
  ) %>% 
  filter(sample_type == measure_type) %>% 
  select(-measure_type) %>% 

  # transform back to wide format
  pivot_wider(names_from = measure, values_from = value) %>% 
  
  # combine with similarly sample-type-separated callables file
  full_join(
    all_expressed_callables %>% 
      rename(all_names = gene_name) %>% 
      rename(gene_name = actual_gene_name) %>% 
      select(-all_names) %>% 
      mutate(
        sample_group = str_extract(sample, '^[A-Z]{2}'),
        sample_type = str_extract(sample, '[05]$')
      ) %>% 
      group_by(sample_group, sample_type, gene_name) %>% 
      summarise(nonsyn = sum(nonsyn), syn = sum(syn)),
    by = c('sample_group', 'sample_type', 'gene_name')
  ) %>% 

  # some data cleanup post full join
  distinct() %>% # full join induced some duplicates
  arrange(sample) %>% 
  mutate(nonsyn_muts = ifelse(is.na(nonsyn_muts), 0, nonsyn_muts),
         syn_muts = ifelse(is.na(syn_muts), 0, syn_muts)) %>% 
  
  # group by sample group and sample type and get sums for mut counts and callable counts
  select(sample, sample_group, sample_type, gene_name, everything()) %>% 
  group_by(sample_group, sample_type) %>% 
  summarise(across(contains('syn'), sum)) %>% 
  
  # compute Ka, Ks, and Ka/Ks values for 0 and 5 samples across sample groups
  mutate(
    ka = nonsyn_muts / nonsyn,
    ks = syn_muts / syn
  ) %>% 
  mutate(
    ka_ks = ka / ks
  ) %>% 
  select(
    sample_group, sample_type, ka_ks, everything()
  )

expressed_ka_ks
```


## Bootstrap confidence intervals

Need to bootstrap original mutation dataset once again, but it'll have to be
exclusively for mutations in expressed genes:

```{r}
muts_described_expressed = read_tsv(
  here('data/mutations/mut_describer/muts_described.final.tsv'),
  col_types = cols()
) %>% 
  
  # keep cols of interest - including feature_ID
  select(
    chromosome, position, sample = mutant_sample, 
    nonsyn_v_syn, gene_name = feature_ID
  ) %>% 
  filter(!is.na(nonsyn_v_syn)) %>% 
  
  # join with list of genes
  inner_join(
    all_expressed, by = c('sample', 'gene_name')
  ) %>% 
  
  # add sample group and type as cols
  mutate(
    sample_group = str_extract(sample, '^[A-Z]{2}'),
    sample_type = str_extract(sample, '[05]$')
  ) %>% 
  select(
    sample_group, sample_type, chromosome, position, sample, gene_name, nonsyn_v_syn
  ) %>% 
  arrange(sample_group, sample_type)
```

### Resampling muts with replacement

```{r}
ka_ks_gene_boot = function(d, indices) {
  # compute ka/ks value for each
  d_resampled = d[indices,]
  d_summarised = d_resampled %>% 
    
    # restructure resampled dataset to match join from earlier
    group_by(sample, gene_name, sample_group, sample_type) %>% 
    count(nonsyn_v_syn) %>% 
    ungroup() %>% 
    pivot_wider(names_from = nonsyn_v_syn, values_from = n, values_fill = 0) 
  
  if (!'synonymous' %in% colnames(d_summarised)) {
    d_summarised$synonymous = 0
  }
  
  # get sample group and type
  current_sample_group = unique(d_summarised$sample_group)
  current_sample_type = unique(d_summarised$sample_type)
  
  # get callables
  d_summarised = d_summarised %>% 
    rename(nonsyn_muts = nonsynonymous, syn_muts = synonymous) %>% 
    
    # get callable counts based on genes
    full_join(
      all_expressed_callables %>% 
        rename(all_names = gene_name) %>% 
        rename(gene_name = actual_gene_name) %>% 
        select(-all_names) %>% 
        mutate(
          sample_group = str_extract(sample, '^[A-Z]{2}'),
          sample_type = str_extract(sample, '[05]$')
        ) %>% 
        filter(
          sample_group == current_sample_group,
          sample_type == current_sample_type
        ) %>% 
        group_by(sample_group, sample_type, gene_name) %>% 
        summarise(nonsyn = sum(nonsyn), syn = sum(syn), .groups = 'drop_last'),
      by = c('sample_group', 'sample_type', 'gene_name')
    ) %>% 
  
    # some data cleanup post full join
    distinct() %>% # full join induced some duplicates
    arrange(sample) %>% 
    replace_na(list(nonsyn_muts = 0, syn_muts = 0)) %>% 
    
    # group by sample group and sample type and get sums for mut counts and callable counts
    select(sample, sample_group, sample_type, gene_name, everything()) %>% 
    group_by(sample_group, sample_type) %>% 
    summarise(across(contains('syn'), sum), .groups = 'drop_last') %>% 
    
    # compute Ka, Ks, and Ka/Ks values for 0 and 5 samples across sample groups
    mutate(
      ka = nonsyn_muts / nonsyn,
      ks = syn_muts / syn
    ) %>% 
    mutate(
      ka_ks = ka / ks
    ) %>% 
    select(
      sample_group, sample_type, ka_ks, everything()
    )
  
  return(d_summarised$ka_ks)
      
}

```

### Calculating confidence intervals

```{r}
get_expressed_cis = function(d, rep_count) {
  tic(msg = paste('bootstrap for', unique(d$sample_factor)))
  message(paste('starting', unique(d$sample_factor)))
  boot_obj = boot(data = d, statistic = ka_ks_gene_boot, R = rep_count)
  ci_obj = boot.ci(boot_obj, type = 'bca')$bca[c(4,5)] # lower and upper bounds
  out = data.frame(
    sample_group = unique(d$sample_group),
    sample_type = unique(d$sample_type),
    lower_ci = ci_obj[1],
    upper_ci = ci_obj[2]
  )
  toc(log = TRUE)
  return(out)
}

expressed_set_cis = muts_described_expressed %>% 
  mutate(sample_factor = paste(sample_group, sample_type, sep = '_')) %>% 
  filter(sample_group != 'SL') %>% 
  split(.$sample_factor) %>% 
  map_dfr(~ get_expressed_cis(., 1000)
  ) # takes a minute per group

expressed_set_cis

```
Joining with orig Ka/Ks values:

```{r}
ka_ks_expressed_final = expressed_ka_ks %>% 
  full_join(expressed_set_cis, by = c('sample_group', 'sample_type')) %>% 
  select(sample_group, sample_type, ka_ks, contains('ci'), everything())

ka_ks_expressed_final

write_tsv(
  ka_ks_expressed_final, 
  file = here('data/rate/ka_ks/ka_ks_expressed_final_CIs.tsv'))
```


# Plots

Ka/Ks for the genome:

```{r}
ka_ks_genome_final %>% 
  ggplot(aes(x = sample_group, y = ka_ks, fill = sample_type)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci, group = sample_type),
    position = position_dodge(width = 0.9),
    size = 0.5, width = 0.25) +
  theme_classic() +
  scale_fill_viridis_d(begin = 0.5, end = 1)
```

For the expressed gene set:

```{r}
ka_ks_expressed_final %>% 
  ggplot(aes(x = sample_group, y = ka_ks, fill = sample_type)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci, group = sample_type),
    position = position_dodge(width = 0.9),
    size = 0.5, width = 0.25) +
  theme_classic() +
  scale_fill_viridis_d(begin = 0.5, end = 1)

```



















