---
title: "Shortest path analysis"
author: Ahmed Hasan
output: pdf_notebook
---

# Package import

```{r}
library(tidyverse)
library(here)
library(fs)
library(boot)
library(tictoc)
library(patchwork)
library(wesanderson)
```

# Data import

```{r}
ma_orig = read_tsv(
  here('data/network/matrices/MA_orig_light.tsv'),
  col_types = cols()
) %>% 
  mutate(type = 'ma')

adaptation = read_tsv(
  here('data/network/matrices/adaptation.tsv'),
  col_types = cols()
) %>% 
  mutate(type = 'adaptation')

salt = read_tsv(
  here('data/network/matrices/saltMA_5_replacement.tsv'),
  col_types = cols()
) %>% 
  mutate(type = 'saltMA')
```

Get min paths, combine with mut describer output, then set shortest paths of repeat genes to 1:

```{r}
adaptation_muts = read_tsv(
  here('data/rate/mut_describer/adaptation.gene_sets.filtered.tsv'),
  col_types = cols()
) %>% 
  filter(genic == 1) %>% 
  select(gene = Gene.primaryIdentifier, chromosome) %>% 
  left_join(
    adaptation %>% 
      bind_rows(adaptation %>% rename(gene1 = gene2, gene2 = gene1)) %>% # double to make sure all genes included
      select(gene = gene1, shortest_path) %>% 
      distinct() %>% 
      group_by(gene) %>% 
      mutate(min_shortest_path = min(shortest_path)) %>% 
      filter(shortest_path == min_shortest_path),
    by = c('gene')
  ) %>% 
  select(-shortest_path) %>% 
  filter(!is.na(min_shortest_path))

ma_muts = read_tsv(
  here('data/rate/mut_describer/ma.gene_sets.tsv'),
  col_types = cols()
) %>% 
  filter(genic == 1) %>% 
  select(chromosome, feature_names) %>% 
  mutate(
    gene = str_match(feature_names, "^\\[\\'([A-Za-z\\.0-9]+)")[, 2] # lol
  ) %>% 
  select(chromosome, gene) %>% 
  left_join(
    ma_orig %>% 
      bind_rows(adaptation %>% rename(gene1 = gene2, gene2 = gene1)) %>% 
      select(gene = gene1, shortest_path) %>% 
      distinct() %>% 
      group_by(gene) %>% 
      mutate(min_shortest_path = min(shortest_path)) %>% 
      filter(shortest_path == min_shortest_path),
    by = c('gene')
  ) %>% 
  select(-shortest_path) %>% 
  filter(!is.na(min_shortest_path))

salt_ma_muts = read_tsv(
  here('data/rate/mut_describer/muts_described.final.gene_sets.tsv'),
  col_types = cols()
) %>% 
  bind_rows(
    read_tsv(
      here('data/rate/mut_describer/indels_described.gene_sets.tsv'), col_types = cols()
    ) %>% 
      mutate(mut_codon = as.character(mut_codon),
             rho = as.character(rho)) # column type incompatibilities
  ) %>% 
  mutate(sample_type = str_extract(mutant_sample, '[05]$')) %>% 
  filter(genic == 1, sample_type == 5) %>% 
  select(chromosome, feature_names) %>% 
  mutate(
    gene = str_match(feature_names, "^\\[\\'([A-Za-z\\.0-9]+)")[, 2] # lol
  ) %>% 
  select(-feature_names) %>% 
  left_join(
    salt %>% 
      bind_rows(adaptation %>% rename(gene1 = gene2, gene2 = gene1)) %>% 
      select(gene = gene1, shortest_path) %>% 
      distinct() %>% 
      group_by(gene) %>% 
      mutate(min_shortest_path = min(shortest_path)) %>% 
      filter(shortest_path == min_shortest_path),
    by = c('gene')
  ) %>% 
  select(-shortest_path) %>% 
  filter(!is.na(min_shortest_path))
  
```

Combining these into lists:

```{r}
networks = list(
  'ma_orig' = ma_orig,
  'adaptation' = adaptation,
  'salt' = salt
) %>% 
  map(
    ~ select(., type, gene1, gene2, shortest_path)
  )

muts_all = list(
  'ma_orig' = ma_muts %>% mutate(type = 'ma'),
  'adaptation' = adaptation_muts %>% mutate(type = 'adaptation'),
  'salt' = salt_ma_muts %>% mutate(type = 'salt')
)
```

Loading in resamples:

```{r}
ma_resamples = read_tsv(
  here('data/network/resamples/ma_resample_100_replacement.tsv'),
  col_types = cols()
)

all_resamples = read_tsv(
  here('data/network/resamples/chlamynet_full_resample_100_replacement.tsv'),
  col_types = cols()
)

salt_ma_resamples = read_tsv(
  here('data/network/resamples/saltMA_resample_100_replacement.tsv'),
  col_types = cols()
)
```

# Data exploration

What are the distributions of shortest paths across these three datasets?

```{r}
# all x all
networks %>% 
  map(
    ~ ggplot(., aes(x = shortest_path)) +
      geom_histogram(binwidth = 1, fill = 'light blue', color = 'black') +
      ggtitle(label = unique(.$type)) +
      coord_cartesian(x = c(0, 20)) +
      theme_classic()
  )

# min paths with number of mutations factored in
muts_all %>% 
  map(
    ~ ggplot(., aes(x = min_shortest_path)) +
      geom_histogram(binwidth = 1, fill = 'steel blue', color = 'black') +
      ggtitle(label = unique(.$type)) +
      coord_cartesian(x = c(0, 20)) +
      theme_classic()
  )
```

Putting these together in a single plot:

```{r}
# all x all
network_means = networks %>% 
  map_dfr(~ group_by(., type) %>% 
            summarise(., mean_shortest_path = mean(shortest_path)))

networks %>% 
  bind_rows() %>% 
  
  ggplot(aes(x = shortest_path)) +
  geom_histogram(
    binwidth = 1,
    fill = 'steelblue',
    color = 'black'
  ) +
  coord_cartesian(
    x = c(0, 20)
  ) +
  facet_wrap(
    ~ type, scales = 'free_y', 
    nrow = 3, ncol = 1) +
  geom_vline(
    data = network_means,
    aes(xintercept = mean_shortest_path),
    size = 1.5,
    color = 'red',
    linetype = 'dotdash'
  )
```

Doing this for the min paths:

```{r}
network_min_means = muts_all %>% 
  map_dfr(~ group_by(., type) %>% 
            summarise(., mean_min_path = mean(min_shortest_path)))

muts_all %>% 
  bind_rows() %>% 
  
  ggplot(aes(x = min_shortest_path)) +
  geom_histogram(
    binwidth = 1,
    fill = 'steelblue',
    color = 'black'
  ) +
  coord_cartesian(
    x = c(0, 20)
  ) +
  facet_wrap(
    ~ type, scales = 'free_y',
    nrow = 3, ncol = 1
  ) +
  geom_vline(
    data = network_min_means,
    aes(xintercept = mean_min_path),
    size = 1.5,
    color = 'red',
    linetype = 'dotdash'
  )
```


Getting some summary statistics:

```{r}
networks %>% 
  map(~ summary(.))
```

## Paring down data to minimum shortest paths

```{r}
networks %>% 
  map(~ group_by(., gene1) %>% 
        mutate(min_shortest_path = min(shortest_path)) %>% 
        filter(shortest_path == min_shortest_path)) %>%  # preserves ties
  bind_rows() %>% 
  
  # plot distributions
  ggplot(aes(x = shortest_path)) +
  geom_histogram(
    binwidth = 1,
    fill = 'steelblue',
    color = 'black'
  ) +
  coord_cartesian(
    x = c(0, 20)
  ) +
  facet_wrap(
    ~ type, scales = 'free_y', 
    nrow = 3, ncol = 1)
```

# Resampling

```{r}
networks_min_paths = networks %>% 
  map(~ group_by(., gene1) %>% 
        mutate(min_shortest_path = min(shortest_path)) %>% 
        filter(shortest_path == min_shortest_path) %>% 
        ungroup() %>% 
        select(type, gene1, min_path = shortest_path))
```

Looking at distribution of MA resamples, stacked

```{r}
# looking at distribution of MA resamples, stacked

ggplot(ma_resamples, aes(x = min_path)) +
  geom_histogram(
    binwidth = 1,
    color = 'black',
    fill = 'steelblue') +
  coord_cartesian(c(0, 20))
```

Plotting side by side with adaptation distribution:


```{r}
palette = wes_palette('Darjeeling1', 5, 'continuous')

resample_dist_plot = bind_rows(
  muts_all$adaptation %>% select(type, min_path = min_shortest_path) %>% mutate(type = 'HS genes'),
  ma_resamples %>% mutate(type = 'MA resamples') %>% select(type, min_path),
  all_resamples %>% mutate(type = 'Random resamples') %>% select(type, min_path)
) %>% 
  mutate(type = fct_relevel(type, c('Random resamples', 'MA resamples', 'HS genes'))) %>% 
  
  # plot distributions
  ggplot(aes(x = min_path - 1, fill = type)) +
  geom_histogram(
    aes(y = ..density..),
    binwidth = 1,
    color = 'black'
  ) +
  coord_cartesian(
    x = c(0, 20)
  ) +
  facet_wrap(
    ~ type, nrow = 3, ncol = 1) +
  theme_classic() +
  labs(
    x = 'Minimum shortest path', 
    y = 'Proportion',
    tag = 'A') +
  theme(
    axis.title = element_text(family = 'Helvetica', size = 12, color = 'black'),
    axis.text = element_text(family = 'Helvetica', size = 12, color = 'black'),
    strip.text = element_text(family = 'Helvetica', size = 12, color = 'black'),
    plot.tag = element_text(family = 'Helvetica', size = 12, face = 'bold'),
    strip.background = element_blank()
  ) +
  scale_x_continuous(breaks = seq(0, 9, 1), labels = seq(0, 9, 1)) +
  coord_cartesian(x = c(0, 9)) +
  guides(fill = FALSE) +
  scale_fill_manual(
    values = c(
      'HS genes' = palette[3],
      'MA resamples' = palette[1],
      'Random resamples' = palette[2]
    )
  )

resample_dist_plot
```

Redoing with position dodge:

```{r}
bind_rows(
  muts_all$adaptation %>% select(type, min_path = min_shortest_path),
  ma_resamples %>% mutate(type = 'MA resamples') %>% select(type, min_path),
  # all_resamples %>% mutate(type = 'chlamynet resamples') %>% select(type, min_path)
) %>% 
  
  # plot distributions
  ggplot(aes(x = min_path, fill = type)) +
  geom_histogram(
    binwidth = 1,
    color = 'black',
    position = 'dodge'
  ) +
  coord_cartesian(
    x = c(0, 20)
  ) +
  theme_classic() +
  labs(x = 'Min shortest path', y = 'Count') +
  theme(
    axis.title = element_text(family = 'Helvetica', size = 12, color = 'black'),
    axis.text = element_text(family = 'Helvetica', size = 12, color = 'black'),
    strip.text = element_text(family = 'Helvetica', size = 12, color = 'black')
  ) +
  scale_x_continuous(breaks = seq(0, 14, 2), labels = seq(0, 14, 2)) +
  coord_cartesian(x = c(0, 14))
```


Looking at a handful of chlamynet resamples:

```{r}
ggplot(ma_resamples %>% filter(iteration < 21), aes(x = min_path)) +
  geom_histogram(
    binwidth = 1,
    fill = 'steelblue',
    color = 'black'
  ) +
  coord_cartesian(
    x = c(0, 15)
  ) +
  facet_wrap(~ iteration)
```

Residual comparisons - 

1. For a given replicate, calculate the difference in bin size for each bin vs adaptation
2. Sum those diffs for an overall distribution difference

```{r}
adaptation_binned = muts_all$adaptation %>% 
  rename(min_path = min_shortest_path) %>% 
  group_by(min_path) %>% 
  summarise(adaptation_count = n())

ma_resamples_binned = ma_resamples %>% 
  group_by(iteration, min_path) %>% 
  summarise(random_count = n(), .groups = 'drop')

# get bin size 
resample_residuals = ma_resamples_binned %>% 
  
  # join adaptation dataset - replace NAs with 0s
  full_join(adaptation_binned, by = c('min_path')) %>% 
  replace_na(list(random_count = 0, adaptation_count = 0)) %>% 
  
  # get differences and return a sum for each iteration
  mutate(sq_diff = (adaptation_count - random_count)) %>% 
  filter(!is.na(iteration))


resample_residuals_summed = resample_residuals %>% 
  group_by(iteration, min_path) %>% # group by iteration as well if doing stacked plot, remove if not
  summarise(mean_sq_diff = mean(sq_diff), 
            sum_sq_diff = sum(sq_diff),
            min_sq_diff = min(sq_diff), 
            max_sq_diff = max(sq_diff))
  
# plot
residual_plot = resample_residuals_summed %>% 
  ggplot(aes(x = min_path - 1, y = sum_sq_diff)) +
  geom_bar(
    stat = 'identity',
    fill = 'steelblue'
  ) +
  # geom_errorbar(
    # aes(ymin = min_sq_diff, ymax = max_sq_diff, width = 0.4)
  # ) +
  scale_x_continuous(breaks = seq(0, 10, 1)) +
  scale_y_continuous(breaks = seq(-2000, 1500, 500)) +
  theme_classic() +
  labs(
    x = 'Minimum shortest path', 
    y = 'HS gene count - resample count',
    tag = 'B') +
  theme(
    axis.title = element_text(family = 'Helvetica', size = 12, color = 'black'),
    axis.text = element_text(family = 'Helvetica', size = 12, color = 'black'),
    plot.tag = element_text(family = 'Helvetica', size = 12, face = 'bold')
  ) +
  geom_hline(yintercept = 0, linetype = 'dashed')

residual_plot
```

Combine plots:

```{r}
network_fig = resample_dist_plot + residual_plot

network_fig

ggsave(here('plots/network.eps'), plot = network_fig, width = 6, height = 4)
```

# Comparing across multiple datasets

## MA vs random

```{r}
all_resamples_combined = all_resamples %>% 
  group_by(min_path) %>% 
  summarise(random_count = n(), .groups = 'drop')

ma_resamples_combined = ma_resamples %>% 
  group_by(min_path) %>% 
  summarise(random_count = n(), .groups = 'drop')

bind_rows(
  all_resamples_combined %>% mutate(type = 'all'),
  ma_resamples_combined %>% mutate(type = 'ma')
) %>% 
ggplot(., aes(x = min_path, y = random_count, fill = type)) +
  geom_bar(stat = 'identity') +
  facet_wrap(~ type)
  
```

Are they statistically different?

```{r}
wilcox.test(all_resamples$min_path, ma_resamples$min_path, alternative = 'two.sided')
```

## MA vs HS

Testing individual MA resamples vs adaptation:

```{r}
wilcox.test(
  ma_resamples %>% 
    filter(iteration == 1) %>% 
    select(min_path) %>% 
    unlist(),
  muts_all$adaptation$min_shortest_path
)
```

```{r}
indiv_test = function(ma, hs, iteration_value) {
  wilcox.test(
    ma %>% 
      filter(iteration == iteration_value) %>% 
      select(min_path) %>% 
      unlist(),
    hs$min_shortest_path,
    alternative = 'two.sided'
  ) %>% 
    broom::tidy()
}

map_dfr(
  c(1:100), ~ indiv_test(ma_resamples, muts_all$adaptation, .), .id = 'iteration'
) %>% 
  summary()
```

## HS vs Salt MA

```{r}
map_dfr(
  c(1:100), ~ indiv_test(salt_ma_resamples, muts_all$adaptation, .), .id = 'iteration'
) %>% 
  summary()
```

## HS vs random

```{r}
map_dfr(
  c(1:100), ~ indiv_test(all_resamples, muts_all$adaptation, .), .id = 'iteration'
) %>% 
  summary()
```

