---
title: "GQ threshold analysis"
author: Ahmed Hasan
output: html_notebook
---

to do:
- examine distribution of GQ scores across records + files

```{r}
library(tidyverse)
library(here)
```

## Loading in data

```{r}
d <- read_csv(
  here('data/mutations/gq_tests/all_candidate/gq_all_samples.txt'), col_types = cols())

snps <- d %>% 
  filter(str_detect(gt1, '^[ATCG]/[ATCG]$'),
         str_detect(gt2, '^[ATCG]/[ATCG]$'))

indels <- d %>% 
  anti_join(snps)
```

## Exploring the data

Is there much discrepancy between the GQ values of the two calls at each site?

```{r}
snps %>% 
  ggplot(aes(x = gt1_qual, y = gt2_qual)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  theme_bw()
```

```{r}
cor.test(snps$gt1_qual, snps$gt2_qual)
```

How do the GQ scores relate to QUAL at each site?

```{r}
snps %>% 
  select(gt1_qual, gt2_qual, qual) %>% 
  gather(site, gq, -qual) %>% 
  ggplot(aes(x = gq, y = qual)) +
  geom_point() +
  facet_wrap(~ site) +
  theme_bw()
```

## Assigning bins

Going to go with the minimum GQ at each site to classify sites - can't use `min`
here because it doesn't operate rowwise (and I don't want to use `rowwise` cause
of how slow and janky it is)

```{r}
min_score_distribution <- snps %>% 
  mutate(min_GQ = case_when(
    gt1_qual < gt2_qual ~ gt1_qual,
    gt1_qual > gt2_qual ~ gt2_qual,
    gt1_qual == gt2_qual ~ gt1_qual
  )) %>% 
  group_by(min_GQ) %>% 
  summarise(n = n(), mean_gt1_depth = mean(gt1_depth),
            mean_gt2_depth = mean(gt2_depth))

# View(min_score_distribution)

ggplot(min_score_distribution, aes(x = min_GQ, y = n)) +
  geom_bar(stat = 'identity', color = 'black', fill = 'grey') +
  coord_cartesian(y = c(0, 100)) +
  theme_classic() +
  geom_vline(xintercept = c(20, 30), linetype = 'dashed')
```

Making a table:

```{r}
min_score_distribution %>% 
  mutate(GQ_bin = case_when(
    min_GQ < 10 ~ '0-9',
    min_GQ >= 10  & min_GQ <= 14 ~ '10-19',
    min_GQ >= 15  & min_GQ < 20 ~ '10-19',
    min_GQ >= 20 & min_GQ <= 24 ~ '20-24',
    min_GQ >= 25 & min_GQ <= 29 ~ '25-29',
    min_GQ >= 30 ~ '30-100'
  )) %>% 
  group_by(GQ_bin) %>% 
  summarise(
    sites = sum(n), 
    min_gt1_depth = round(min(mean_gt1_depth), 2),
    mean_gt1_depth = round(mean(mean_gt1_depth), 2),
    min_gt2_depth = round(min(mean_gt2_depth), 2),
    mean_gt2_depth = round(mean(mean_gt2_depth), 2))
```


and a cdf:

```{r}
snps %>% 
  mutate(min_GQ = case_when(
    gt1_qual < gt2_qual ~ gt1_qual,
    gt1_qual > gt2_qual ~ gt2_qual,
    gt1_qual == gt2_qual ~ gt1_qual
  )) %>% 
  ggplot(aes(x = 1 - min_GQ)) +
  stat_ecdf() +
  theme_bw() +
  geom_vline(xintercept = c(-20, -30), linetype = 'dashed') +
  scale_x_continuous(
    breaks = seq(-100, 0, 10), 
    labels = seq(100, 0, -10)) +
  labs(
    x = 'min GQ',
    y = 'proportion'
  ) +
  coord_cartesian(x = c(-40, -10), y = c(0, 0.01))
```

## By sample

```{r}
by_sample <- snps %>% 
  mutate(min_GQ = case_when(
    gt1_qual < gt2_qual ~ gt1_qual,
    gt1_qual > gt2_qual ~ gt2_qual,
    gt1_qual == gt2_qual ~ gt1_qual
  )) %>% 
  group_by(fname, min_GQ) %>% 
  summarise(n = n(), mean_gt1_depth = mean(gt1_depth),
            mean_gt2_depth = mean(gt2_depth)) %>% 
  mutate(GQ_bin = case_when(
    min_GQ < 10 ~ '0-9',
    min_GQ >= 10  & min_GQ < 20 ~ '10-19',
    min_GQ >= 20 & min_GQ <= 24 ~ '20-24',
    min_GQ >= 25 & min_GQ <= 29 ~ '25-29',
    min_GQ >= 30 ~ '30-100'
  )) %>% 
  group_by(fname, GQ_bin) %>% 
  summarise(
    sites = sum(n))

by_sample
```

Visualizing:

```{r}
by_sample %>% 
  ggplot(aes(x = GQ_bin, y = sites)) +
  geom_bar(stat = 'identity') +
  facet_wrap(~ fname) + 
  coord_cartesian(y = c(0, 200)) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))
```












